### Dataset similarity

#### By gene standard deviation

```{r}
sd_dat <- gene_weights %>%
  select(matches('_sd$'))

sd_cor_mat <- cor(sd_dat, use = 'pairwise.complete.obs')

heatmaply(sd_cor_mat)
```

##### By gene MAD

```{r}
mad_dat <- gene_weights %>%
  select(matches('_mad$'))

mad_cor_mat <- cor(mad_dat, use = 'pairwise.complete.obs')

heatmaply(mad_cor_mat)
```

##### By gene dip statistic

```{r}
dip_dat <- gene_weights %>%
  select(matches('_dip_stat$'))

dip_cor_mat <- cor(dip_dat, use = 'pairwise.complete.obs')

heatmaply(dip_cor_mat)
```

#### Missing data

```{r, fig.height = 24, fig.width = 8}
plot_missing(gene_weights, ggtheme = theme_bw_high_res())
```

### Data distributions

#### Standard Deviation

```{r fig.width = 12, fig.height = 6}
ind <- grepl('_sd', colnames(gene_weights))

dat <- gene_weights[, ind] %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

upper_lim <- quantile(dat$statistic, 0.9)

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  facet_wrap(~dataset) +
  xlim(0, upper_lim) +
  ylim(0, 1) +
  xlab("Standard Deviation (Clipped)") +
  theme_bw_high_res()
```

#### Median Absolute Deviation

```{r fig.width = 12, fig.height = 6}
ind <- grepl('_mad', colnames(gene_weights))

dat <- gene_weights[, ind] %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

upper_lim <- quantile(dat$statistic, 0.9)

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  facet_wrap(~dataset) +
  xlim(0, upper_lim) +
  ylim(0, 1) +
  xlab("MAD (Clipped)") +
  theme_bw_high_res()
```

#### Kurtosis

```{r fig.width = 12, fig.height = 6}
ind <- grepl('_kurtosis', colnames(gene_weights))

dat <- gene_weights[, ind] %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

upper_lim <- quantile(dat$statistic, 0.8)

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  facet_wrap(~dataset) +
  xlim(0, upper_lim) +
  ylim(0, 0.5) +
  xlab("Kurtosis (Clipped)") +
  theme_bw_high_res()
```

#### Hartigan's Dip Test

##### Dip Stat

```{r fig.width = 12, fig.height = 6}
ind <- grepl('_dip_stat', colnames(gene_weights))

dat <- gene_weights[, ind] %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  facet_wrap(~dataset, scales = 'free') +
  xlab("Dip Stat") +
  theme_bw_high_res()
```

##### Dip P-value

```{r fig.width = 12, fig.height = 6}
ind <- grepl('_dip_pval', colnames(gene_weights))

dat <- gene_weights[, ind] %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  facet_wrap(~dataset, scales = 'free') +
  xlab("Dip P-value") +
  theme_bw_high_res()
```

#### Multiple Myeloma Disease Stage

```{r fig.width = 12, fig.height = 6}
ind <- grepl('_stage', colnames(gene_weights))

dat <- gene_weights[, ind] %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  xlim(0, 1) +
  facet_wrap(~dataset, scales = 'free_y') +
  xlab("R^2 (Disease Stage)") +
  theme_bw_high_res()
```

#### Multiple Myeloma Status

```{r fig.width = 3, fig.height = 3}
ind <- grepl('_status', colnames(gene_weights))

dat <- gene_weights[, ind, drop = FALSE] %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  xlim(0, 1) +
  xlab("R^2 (Disease Status)") +
  theme_bw_high_res()
```

#### Multiple Myeloma Treatment Response

```{r fig.width = 12, fig.height = 6}
ind <- grepl('_response', colnames(gene_weights))

dat <- gene_weights[, ind] %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  xlim(0, 1) +
  facet_wrap(~dataset, scales = 'free_y') +
  xlab("R^2 (Treatment Response)") +
  theme_bw_high_res()
```

#### Multiple Myeloma Survival (Cox statistic)

```{r fig.width = 12, fig.height = 6}
ind <- grepl('_cox_stat', colnames(gene_weights))

dat <- gene_weights[, ind] %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  xlim(0, 30) +
  facet_wrap(~dataset, scales = 'free_y') +
  xlab("Cox statistic (Clipped)") +
  theme_bw_high_res()
```

#### Multiple Myeloma Survival (Cox P-value)

```{r fig.width = 12, fig.height = 6}
ind <- grepl('_cox_pval', colnames(gene_weights))

dat <- gene_weights[, ind] %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  facet_wrap(~dataset, scales = 'free_y') +
  xlab("Cox P-value") +
  theme_bw_high_res()
```
