## Dataset similarity

### By gene expression

#### Gene expression overview

Below, gene expression heatmaps are shown for each individual dataset for a subsample of 
500 randomly selected genes. Genes are chosen such that they are measured in across all
datasets.

```{r measure_missing_data}
# number of missing values for each gene
num_na <- apply(gene_weights[, -1], 1, function(x) { sum(is.na(x)) })

#table(num_na == 0)
# 
# FALSE  TRUE 
# 56469  5300 

# genes measured across all datasets
complete_genes <- gene_weights$gene_symbol[num_na == 0]
```

```{r indiv_gene_expr_heatmap, results = 'asis'}
# subset of n = 500 genes with the largest variance across all datasets
gene_vars <- gene_stats %>%
  select(gene_symbol, ends_with('_var')) %>%
  filter(gene_symbol %in% complete_genes) %>%
  column_to_rownames('gene_symbol')

gene_var_totals <- rowSums(gene_vars)

max_var_genes <- names(head(sort(gene_var_totals, decreasing = TRUE), 500))

# order genes based on the average expression for each dataset
dat <- gene_stats %>%
  filter(gene_symbol %in% max_var_genes) %>%
  select(gene_symbol, ends_with('_mean')) %>%
  column_to_rownames('gene_symbol')

# determine gene ordering to use
gene_order <- hclust(dist(dat))$order

# next, load each individual expression dataset and generate a heatmap for the
# specified genes, keeping the order of the genes constant.

# gene expression datasets
infiles <- Sys.glob(file.path(geo_dir, '*', 'processed', sprintf("*_gene_expr.csv")))

for (infile in infiles) {
  # accession
  acc <- sub('_gene_expr.csv', '', basename(infile))

  cat(sprintf('#### %s\n', acc))

  # load expression data, collapse multi-mapped genes, and select subset genes
  dat <- read_csv(infile, col_types = cols()) %>%
    select(-matches('ensgene'), -matches('probe_id')) %>%
    separate_rows(gene_symbol, sep = " ?//+ ?") %>%
    group_by(gene_symbol) %>%
    filter(!is.na(gene_symbol)) %>%
    summarize_all(sum) %>%
    filter(gene_symbol %in% max_var_genes) %>%
    column_to_rownames('gene_symbol')

  # plot heatmap
  aheatmap(round(log2(dat + 1), 3), color = viridis(100), Rowv = FALSE, main = acc)
}
```

#### Gene expression correlation heatmap

Next, average expression for the ~5,300 genes that are measured across all datasets 
is compared across datasets. The total average expression within each dataset is
normalized 

```{r, fig.width = 6, fig.height = 4}
# get average expression for each dataset
dat <- gene_stats %>%
  select(gene_symbol, ends_with('mean')) %>%
  column_to_rownames('gene_symbol')

colnames(dat) <- sub('_mean', '', colnames(dat))

# limit to genes measured in all datasets
dat <- dat[rownames(dat) %in% complete_genes, ]

# normalize totals across datasets
dat <- sweep(dat, 2, colSums(dat), '/') * 1E6

cor_mat <- round(cor(dat), 2)

# dataset annotations
annot_df <- datasets[match(colnames(dat), datasets$accession), ] %>%
  select(disease_stage, platform) %>%
  as.data.frame

annot_df$stage <- rep('Mixed', nrow(cor_mat))
annot_df$stage[annot_df$disease_stage == 'Newly Diagnosed'] <- 'Newly Diagnosed'
annot_df$stage[annot_df$disease_stage == 'Relapsed'] <- 'Relapsed'

annot_df <- annot_df %>%
  select(-disease_stage)

dnames <- datasets$dataset[match(colnames(cor_mat), datasets$accession)]
colnames(cor_mat) <- dnames
rownames(cor_mat) <- dnames

aheatmap(cor_mat, color = viridis(100), Colv = FALSE, annRow = annot_df, annColors =
         c('Dark2', 'Set2'))
```

### By gene standard deviation

Comparison of gene standard deviations across datasets.

```{r sd_cor_mat}
sd_dat <- gene_stats %>%
  select(matches('_sd$'))

sd_cor_mat <- round(cor(sd_dat, use = 'pairwise.complete.obs'), 2)

heatmaply(sd_cor_mat)
```

#### By gene MAD

Comparison of gene median absolute deviations across datasets.

```{r mad_cor_mat}
mad_dat <- gene_stats %>%
  select(matches('_mad$'))

mad_cor_mat <- round(cor(mad_dat, use = 'pairwise.complete.obs'), 2)

heatmaply(mad_cor_mat)
```

#### By gene dip statistic

Comparison of gene dip statistics across datasets.

```{r dip_cor_mat}
dip_dat <- gene_stats %>%
  select(ends_with('_dip_score'))

# remove inf values (fixed upstream 2019/07/26...)
dip_dat[dip_dat == Inf] <- 10

dip_cor_mat <- round(cor(dip_dat, use = 'pairwise.complete.obs'), 2)

heatmaply(dip_cor_mat)
```

## Data distributions

### Standard Deviation

```{r fig.width = 12, fig.height = 6}
dat <- gene_stats %>%
  select(ends_with('_sd')) %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

upper_lim <- quantile(dat$statistic, 0.9)

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  facet_wrap(~dataset) +
  xlim(0, upper_lim) +
  ylim(0, 1) +
  xlab("Standard Deviation (Clipped)") +
  theme_bw_high_res()
```

### Median Absolute Deviation

```{r fig.width = 12, fig.height = 6}
dat <- gene_stats %>%
  select(ends_with('_mad')) %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

upper_lim <- quantile(dat$statistic, 0.9)

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  facet_wrap(~dataset) +
  xlim(0, upper_lim) +
  ylim(0, 1) +
  xlab("MAD (Clipped)") +
  theme_bw_high_res()
```

### Kurtosis

```{r fig.width = 12, fig.height = 6}
dat <- gene_stats %>%
  select(ends_with('_kurtosis')) %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

upper_lim <- quantile(dat$statistic, 0.8)

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  facet_wrap(~dataset) +
  xlim(0, upper_lim) +
  ylim(0, 0.5) +
  xlab("Kurtosis (Clipped)") +
  theme_bw_high_res()
```

### Hartigan's Dip Test

#### Dip Stat

```{r fig.width = 12, fig.height = 6}
dat <- gene_stats %>%
  select(ends_with('_dip_score')) %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  facet_wrap(~dataset, scales = 'free') +
  xlab("Dip Stat") +
  theme_bw_high_res()
```

### Covariate P-values

```{r fig.width = 12, fig.height = 6}
dat <- gene_weights %>%
  select(ends_with('_pval')) %>%
  gather(key = covariate, value = pvalue) %>%
  na.omit

ggplot(dat, aes(pvalue, group = covariate, fill = covariate, label = covariate)) +
  geom_density() +
  xlim(0, 1) +
  facet_wrap(~covariate, scales = 'free_y') +
  xlab("P-value") +
  theme_bw_high_res()
```

