---
title: Multiple Myeloma Dataset Comparison
author: V. Keith Hughitt
date: "`r format(Sys.time(), '%d %B, %Y') `"
output:
  html_document:
    toc: true
    number_sections: true
---

# TODO

- Generate "thumbnail" overview images of each dataset.
- Compute correlation of expression data (e.g. sample 1000 random
  genes and measure correlation of expr across datasets)
- Annotate datasets?
  - Are there any obvious factors that could explain why some datasets are more similar
    to others?
    - disease stage (MGUS vs. MM)
    - cell line vs. patient
    - dataset size
    - response measure type
    - platform
- Visualize some examples at each step? (e.g. compare count distributions for a single
  gene with high dip test?)
- For categorical covariates, use limma/deseq to perform differential expression
  for each pair of response levels and record gene statistics / enrichment; provides
  an alternative to the current logistic regression based approach.

# IDEAS

1. For final gene weights, use mostly survival, etc. R^2's since these are likely to be
   the genes that are particularly important for MM.
2. For sd, etc. could try downloading a large number of "normal" (non-MM /
   non-cancer) samples and measure the same statistics. Then look for genes which have
   the highest sd in the MM samples, relative to healthy ones?
3. Use RCellMiner, etc. to select drugs with various mechanisms and generate
   gene weights targetted at each different type of drug mechanism?
   
**Related work:**

1. [Botta et al. (2016)](https://www.nature.com/articles/bcj2016118) - Combines data
   from 5 different MM experiments (table S1) to determine a survival-specific gene signature.
   - focussed on inflammation-related gene subset.

```{r, message = FALSE}
library(annotables)
library(DataExplorer)
library(inspectdf)
library(diptest)
library(e1071)
library(GSEABase)
library(fgsea)
library(heatmaply)
library(knitr)
library(parallel)
library(samr)
library(survival)
library(tidyverse)
```

```{r child='rmd/setup.Rmd'}
```

```{r rmarkdown_settings, include = FALSE}
# stop on encountered an error in a child rmd file
opts_chunk$set(error = FALSE)
opts_knit$set(verbose = TRUE)
```

```{r settings}
# maximum ratio of covariate variables to allow before exluding a gene
# from the final score computation
GENE_MAX_MISSING_COVARIATES <- 0.5
```

## Methods

```{r child='rmd/dataset_summary.Rmd'}
```

```{r child='rmd/load_gene_sets.Rmd'}
```

```{r child='rmd/load_geo.Rmd'}
```

```{r child='rmd/load_mmrf.Rmd'}
```

```{r child='rmd/load_gene_weights.Rmd'}
```

## Results

### Dataset similarity

#### By gene standard deviation

```{r}
sd_dat <- gene_weights %>%
  select(matches('_sd$'))

sd_cor_mat <- cor(sd_dat, use = 'pairwise.complete.obs')

heatmaply(sd_cor_mat)
```

##### By gene MAD

```{r}
mad_dat <- gene_weights %>%
  select(matches('_mad$'))

mad_cor_mat <- cor(mad_dat, use = 'pairwise.complete.obs')

heatmaply(mad_cor_mat)
```

##### By gene dip statistic

```{r}
dip_dat <- gene_weights %>%
  select(matches('_dip_stat$'))

dip_cor_mat <- cor(dip_dat, use = 'pairwise.complete.obs')

heatmaply(dip_cor_mat)
```

##### By covariate weights

```{r}
covariate_dat <- gene_weights %>%
  select(-gene_symbol, -ensgene, 
         -matches('_mean$'), -matches('_median$'), -matches('_sd$'),
         -matches('_mad$'), -matches('_kurtosis$'),
         -matches('_dip_stat$'), -matches('_dip_pval$'))

rownames(covariate_dat) <- gene_weights$gene_symbol

# number of missing entries for each gene
gene_num_na <- apply(covariate_dat, 1, function(x) { sum(is.na(x)) })

#table(gene_num_na)
# gene_num_na
#     0     1     2     3     4     5     6     7     8     9    10    11    12    13    14    15 
#  5465   206   176  6250   571   672  2676   369   810   834   910   526   335  1604    38   168 
#    16    17    18    19    20    21    22    23    24    25    26    27    28 
#   106  1192    52     8    93   578   442   954 19075  1702 14054   977   632 

max_na <- ceiling(GENE_MAX_MISSING_COVARIATES * ncol(covariate_dat))
covariate_dat <- covariate_dat[gene_num_na <= max_na, ]

covariate_cor_mat <- round(cor(covariate_dat, use = 'pairwise.complete.obs'), 2)

heatmaply(covariate_cor_mat, draw_cellnote = TRUE, cellnote_size = 8)
```


#### Missing data

```{r, fig.height = 24, fig.width = 8}
plot_missing(gene_weights, ggtheme = theme_bw_high_res())
```

#### Data distributions

##### Standard Deviation

```{r fig.width = 12, fig.height = 6}
ind <- grepl('_sd', colnames(gene_weights))

dat <- gene_weights[, ind] %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

upper_lim <- quantile(dat$statistic, 0.9)

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  facet_wrap(~dataset) +
  xlim(0, upper_lim) +
  ylim(0, 1) +
  xlab("Standard Deviation (Clipped)") +
  theme_bw_high_res()
```

##### Median Absolute Deviation

```{r fig.width = 12, fig.height = 6}
ind <- grepl('_mad', colnames(gene_weights))

dat <- gene_weights[, ind] %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

upper_lim <- quantile(dat$statistic, 0.9)

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  facet_wrap(~dataset) +
  xlim(0, upper_lim) +
  ylim(0, 1) +
  xlab("MAD (Clipped)") +
  theme_bw_high_res()
```

##### Kurtosis

```{r fig.width = 12, fig.height = 6}
ind <- grepl('_kurtosis', colnames(gene_weights))

dat <- gene_weights[, ind] %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

upper_lim <- quantile(dat$statistic, 0.8)

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  facet_wrap(~dataset) +
  xlim(0, upper_lim) +
  ylim(0, 0.5) +
  xlab("Kurtosis (Clipped)") +
  theme_bw_high_res()
```

##### Hartigan's Dip Test

###### Dip Stat

```{r fig.width = 12, fig.height = 6}
ind <- grepl('_dip_stat', colnames(gene_weights))

dat <- gene_weights[, ind] %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  facet_wrap(~dataset, scales = 'free') +
  xlab("Dip Stat") +
  theme_bw_high_res()
```

###### Dip P-value

```{r fig.width = 12, fig.height = 6}
ind <- grepl('_dip_pval', colnames(gene_weights))

dat <- gene_weights[, ind] %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  facet_wrap(~dataset, scales = 'free') +
  xlab("Dip P-value") +
  theme_bw_high_res()
```

##### Multiple Myeloma Disease Stage

```{r fig.width = 12, fig.height = 6}
ind <- grepl('_stage', colnames(gene_weights))

dat <- gene_weights[, ind] %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  xlim(0, 1) +
  facet_wrap(~dataset, scales = 'free_y') +
  xlab("R^2 (Disease Stage)") +
  theme_bw_high_res()
```

##### Multiple Myeloma Status

```{r fig.width = 3, fig.height = 3}
ind <- grepl('_status', colnames(gene_weights))

dat <- gene_weights[, ind, drop = FALSE] %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  xlim(0, 1) +
  xlab("R^2 (Disease Status)") +
  theme_bw_high_res()
```

##### Multiple Myeloma Treatment Response

```{r fig.width = 12, fig.height = 6}
ind <- grepl('_response', colnames(gene_weights))

dat <- gene_weights[, ind] %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  xlim(0, 1) +
  facet_wrap(~dataset, scales = 'free_y') +
  xlab("R^2 (Treatment Response)") +
  theme_bw_high_res()
```

##### Multiple Myeloma Survival (Cox statistic)

```{r fig.width = 12, fig.height = 6}
ind <- grepl('_cox_stat', colnames(gene_weights))

dat <- gene_weights[, ind] %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  xlim(0, 30) +
  facet_wrap(~dataset, scales = 'free_y') +
  xlab("Cox statistic (Clipped)") +
  theme_bw_high_res()
```

##### Multiple Myeloma Survival (Cox P-value)

```{r fig.width = 12, fig.height = 6}
ind <- grepl('_cox_pval', colnames(gene_weights))

dat <- gene_weights[, ind] %>%
  gather(key = dataset, value = statistic) %>%
  na.omit

ggplot(dat, aes(statistic, group = dataset, fill = dataset, label = dataset)) +
  geom_density() +
  facet_wrap(~dataset, scales = 'free_y') +
  xlab("Cox P-value") +
  theme_bw_high_res()
```

### Functional Enrichment

#### Commonly enriched pathways / gene sets (# datasets)

```{r fgsea_results, message = FALSE, results = 'asis'}
# load fgsea results
fgsea_res_all <- bind_rows(lapply(Sys.glob('output/fgsea_results_*'), read_csv))

fgsea_res_all$dataset <- str_split(fgsea_res_all$covariate, '_', simplify = TRUE)[, 1] 

# number of datasets a pathway was found to be enriched for for at least one
# covariate
fgsea_res_datasets <- fgsea_res_all %>%
  group_by(dataset, gene_set, pathway) %>%
  summarize(mean_padj = mean(padj), median_padj = median(padj)) %>%
  group_by(gene_set, pathway) %>%
  summarize(num_datasets = n(), mean_padj = mean(mean_padj), median_padj = median(median_padj)) %>%
  arrange(desc(num_datasets))

fgsea_res_datasets %>%
  head(20) %>%
  kable
```

#### Commonly enriched pathways / gene sets (# datasets/covariates)

```{r results = 'asis'}
# which pathways are enriched for the most datasets / covariates?
fgsea_res_covariates <- fgsea_res_all %>%
  group_by(gene_set, pathway) %>%
  summarize(num_enriched = n(), mean_padj = mean(padj), median_padj = median(padj)) %>%
  arrange(desc(num_enriched))

fgsea_res_covariates %>%
  head(20) %>%
  kable
```

#### Most significantly enriched pathways / genesets

```{r, results = 'asis'}
# which pathways have the most statistically significant enrichment?
fgsea_res_covariates %>%
  filter(num_enriched >= 3) %>%
  arrange(median_padj) %>%
  head(20) %>%
  kable
```

#### Relationship between covariates and enrichment results

```{r results = 'asis'}
# number of enriched annotations for each covariate assessed
fgsea_summary <- fgsea_res_all %>% 
  group_by(covariate) %>% 
  summarize(num_enriched = n()) %>% 
  arrange(desc(num_enriched))

# statistics / p-values for associated with each fgsea result
covariate_stats <- gene_weights[, fgsea_summary$covariate]

fgsea_summary <- bind_cols(
    fgsea_summary, 
    min_stat    = round(sapply(covariate_stats, min, na.rm = TRUE), 3),
    median_stat = round(sapply(covariate_stats, median, na.rm = TRUE), 3),
    mean_stat   = round(sapply(covariate_stats, mean, na.rm = TRUE), 3),
    max_stat    = round(sapply(covariate_stats, max, na.rm = TRUE), 3))

fgsea_summary$accession <- str_split(fgsea_summary$covariate, '_', simplify = TRUE)[, 1] 

fgsea_summary <- fgsea_summary %>%
  inner_join(datasets %>% select(accession, num_samples, num_genes, year), by = 'accession') %>%
  select(-accession)

fgsea_summary %>%
  kable

# correlation of enrichment with other dataset / covariate features
cor(fgsea_summary[, -1])

# covariates with no associated enrichment
no_sig_results <- colnames(gene_weights)[!colnames(gene_weights) %in% fgsea_summary$covariate]
no_sig_results <- no_sig_results[!grepl('gene_symbol|ensgene|_mean|_median|_sd|_mad|_kurtosis|_stat', no_sig_results)]

no_sig_results
```

#### Most useful gene set collections

```{r}
fgsea_res_collections <- fgsea_res_all %>%
  group_by(gene_set) %>%
  summarize(num_enriched = n(), mean_padj = mean(padj), median_padj = median(padj)) %>%
  rename(collection = gene_set)

# gene set stats
gene_set_stats <- data.frame(collection = names(gene_sets), 
                             num_gene_sets = as.numeric(unlist(lapply(gene_sets, length))))

gene_set_stats$min_size <- unlist(lapply(gene_sets, function(x) { min(unlist(lapply(x, length))) }))
gene_set_stats$max_size <- unlist(lapply(gene_sets, function(x) { max(unlist(lapply(x, length))) }))
gene_set_stats$mean_size <- unlist(lapply(gene_sets, function(x) { mean(unlist(lapply(x, length))) }))
gene_set_stats$median_size <- unlist(lapply(gene_sets, function(x) { median(unlist(lapply(x, length))) }))

gene_set_stats <- gene_set_stats %>%
  inner_join(fgsea_res_collections, by = 'collection') %>%
  mutate(ratio_enriched = num_enriched / num_gene_sets) %>%
  arrange(desc(ratio_enriched))

gene_set_stats %>%
  head(10) %>%
  kable

# is the ratio of enrichment correlated with any properties of the gene set collections?
cor_mat <- gene_set_stats %>%
  select(collection, num_gene_sets, min_size, max_size, mean_size, ratio_enriched) %>%
  column_to_rownames('collection') %>%
  cor 

diag(cor_mat) <- 0

#cor_mat
#                num_gene_sets min_size max_size mean_size ratio_enriched
# num_gene_sets         0.0000  -0.0833   0.6781   -0.0929        -0.1140
# min_size             -0.0833   0.0000  -0.1100    0.6951         0.5800
# max_size              0.6781  -0.1100   0.0000   -0.0444        -0.0639
# mean_size            -0.0929   0.6951  -0.0444    0.0000         0.9171
# ratio_enriched       -0.1140   0.5800  -0.0639    0.9171         0.0000

# larger gene sets are more likely to be enriched..
cor(gene_set_stats$mean_size, gene_set_stats$ratio_enriched)
```

```{r save_fgsea_summary_results}
# save fgsea summary results
write_csv(fgsea_res_all, 'output/fgsea_combined_results_all.csv')
write_csv(fgsea_res_datasets, 'output/fgsea_combined_results_datasets.csv')
write_csv(fgsea_res_covariates, 'output/fgsea_combined_results_covariates.csv')
write_csv(fgsea_res_collections, 'output/fgsea_combined_results_collections.csv')
write_csv(gene_set_stats, 'output/fgsea_combined_results_collections_stats.csv')
```

### Most informative genes

#### All covariates (mean)

```{r, fig.width = 6, fig.height = 6}
# convert p-values to -log10(pvalue) scores so that directionality is the same for
# all variables
covariate_dat_log10pvals <- covariate_dat %>%
  mutate_at(vars(matches("_pval_?\\d*$")), function (x) { -log10(x) }) %>%
  mutate_each(list(scale))


# max-scale and shift to [0, 1]
max_scale <- function(x) {
  x <- x - min(x, na.rm = TRUE)
  x / max(x, na.rm = TRUE)
}
dat <- sapply(covariate_dat_log10pvals, max_scale)
rownames(dat) <- rownames(covariate_dat)

# shift to right so that lowest value for each column is zero
#for (x in colnames(covariate_dat_log10pvals)) {
#  covariate_dat_log10pvals[, x] <- covariate_dat_log10pvals[, x] - min(covariate_dat_log10pvals[, x], na.rm = TRUE)
#}

# next compute average weights for each gene;
# note that some genes are only measured in one or a few of the datasets, so confidence
# of the gene weights should be considered

# how much each column contributes to the score..
#range(colSums(dat, na.rm = TRUE))
# [1]   336 15989

# normalize across columns so that each covariate contributes equally
dat <- sweep(dat, 2, colSums(dat, na.rm = TRUE), '/') * 1E6

ind <- match(rownames(dat), grch37$symbol)

score_dat <- data.frame(
  gene = rownames(dat),
  type = grch37$biotype[ind],
  description = grch37$description[ind],
  mean_score = rowSums(dat, na.rm = TRUE),
  median_score = apply(dat, 1, median, na.rm = TRUE)
)

score_dat %>% 
  arrange(desc(mean_score)) %>%
  head(20) %>%
  kable
```

#### All covariates (median)

```{r results = 'asis'}
score_dat %>% 
  arrange(desc(median_score)) %>%
  head(20) %>%
  kable
```

#### Treatment covariates (mean)

```{r, fig.width = 6, fig.height = 6}
dat <- covariate_dat_log10pvals[, grepl('_response$', colnames(covariate_dat))]

# max-scale and shift to [0, 1]
dat <- sapply(dat, max_scale)
rownames(dat) <- rownames(covariate_dat)

# normalize across columns so that each covariate contributes equally
dat <- sweep(dat, 2, colSums(dat, na.rm = TRUE), '/') * 1E6

ind <- match(rownames(dat), grch37$symbol)

treatment_score_dat <- data.frame(
  gene = rownames(dat),
  type = grch37$biotype[ind],
  description = grch37$description[ind],
  mean_score = rowSums(dat, na.rm = TRUE),
  median_score = apply(dat, 1, median, na.rm = TRUE)
)

treatment_score_dat %>% 
  arrange(desc(mean_score)) %>%
  head(20) %>%
  kable
```

#### Treatment covariates (median)

```{r results = 'asis'}
treatment_score_dat %>% 
  arrange(desc(median_score)) %>%
  head(20) %>%
  kable
```

#### Survival covariates (mean)

```{r, fig.width = 6, fig.height = 6}
dat <- covariate_dat_log10pvals[, grepl('_cox_|_died', colnames(covariate_dat))]

# max-scale and shift to [0, 1]
dat <- sapply(dat, max_scale)
rownames(dat) <- rownames(covariate_dat)

# normalize across columns so that each covariate contributes equally
dat <- sweep(dat, 2, colSums(dat, na.rm = TRUE), '/') * 1E6

ind <- match(rownames(dat), grch37$symbol)

survival_score_dat <- data.frame(
  gene = rownames(dat),
  type = grch37$biotype[ind],
  description = grch37$description[ind],
  mean_score = rowSums(dat, na.rm = TRUE),
  median_score = apply(dat, 1, median, na.rm = TRUE)
)

survival_score_dat %>% 
  arrange(desc(mean_score)) %>%
  head(20) %>%
  kable
```

#### Survival covariates (median)

```{r results = 'asis'}
survival_score_dat %>% 
  arrange(desc(median_score)) %>%
  head(20) %>%
  kable
```

```{r}
stopCluster(cl)
```

#### Functional Enrichment (Covariate Scores)

```{r fgsea, results = 'asis', warnings = FALSE}
gene_stats <- setNames(score_dat$mean_score, score_dat$gene)

for (gs in names(gene_sets)) {
	res <- fgsea(gene_sets[[gs]], gene_stats, nperm = 15000, nproc = MAX_THREADS) %>%
		filter(padj < 0.05) %>%
		select(-leadingEdge) %>%
		arrange(padj)

	if (nrow(res) > 0) {
    cat(sprintf("\n##### %s (n = %d)\n", gs, nrow(res)))
	  print(kable(head(res, 20)))
	}
}
```

```{r}
sessionInfo()
```

